---
title: "Mind the match"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Source the functions and libraries
```{r cars}
source("./source/functions.R")

source("./source/libraries.R")
```

# Matching tool
Did you ever needed to match 100s of samples in different sequencing batches? No? Good... But in case you will ever need to do it this script will help you to get it done properly.

In this script you can input a table *start_data* including the clinical parameters of the samples you want to batch correctly and after defining some parameters you can run an endless loop to get the best combination of samples!

This script is not meant to be strictly reproducible, is more a one time one, if you want to make it reproducible uncomment the set.seed line.

## Definition of the parameters
```{r}
# set.seed(91)

n_groups <- 6 # Number of batches you want to generate
group_size <- 12 # Number of sample in each batch

start_data <- read_xlsx("./data/example_dataset.xlsx")

numeric_to_optimize <- c("Day") # If none is wanted set to NULL
categorical_to_optimize <- c("Grup", "Sex") # If none is wanted set to NULL

n_perm <- Inf
```

## Running the model

### Defining some standard variables
```{r}
# Defining the new groups according to the parameter set
new_groups <- vector()

for (group_num in 1:n_groups) {
  
  new_groups <- c(new_groups, rep(group_num, group_size))
  
  rm(group_num)
  
}

# If the number of samples is not divisible for the group correctly (e.g. the last group os smaller), we need a correction
if (group_size*n_groups != dim(start_data)[1]) {
  new_groups <- new_groups[1:dim(start_data)[1]]
}
```

### Preparing a container for the results
```{r}
# Generate a place to collect the results
permutation <- list()

permutation$orders <- vector() # This is the container for the list order

permutation$reults <- Inf # This is the container for the final score
```

### Permutations
```{r}
# Initialize a counter
counter <- 0
permut <- 0

repeat {
  
  # Add a counter for the permuation - Just to see how fast it goes
  permut <- permut + 1
  
  if (permut %% 10000 == 0) {
    print(paste("Permutation:", permut, sep = " "))
  }
  
  if (permut == n_perm){ # This is because is not possible to set a for loop to infinite, in this case I need the repeat function with a breack rule
    break
  }
  
  # Generate a new random order of the data
  random_order <- sample(seq(1:dim(start_data)[1]), dim(start_data), replace = FALSE)

  # Adding the new groups to the data table with the new order
  new_data <- start_data[random_order, ]
  new_data$new_groups <- new_groups # This is the new grouping which we should test

  # From here I can add the rules
  result_int <- 0
  
  ## Optimization of numeric
  if (!is.null(numeric_to_optimize)) {
    
    for (variable_num in numeric_to_optimize) {
    
      result_int <- result_int + weight_numeric(variable_num)
    
    }
    
  }
  
  ## Optimization of categorical
  if (!is.null(categorical_to_optimize)) {
    
    for (variable_cat in categorical_to_optimize) {
    
      result_int <- result_int + weight_categoric(variable_cat)
    
    }
    
  }
  
  # If the result is better save it!
  if (result_int < permutation$reults) {
    
    permutation$orders <- random_order
    permutation$reults <- result_int
    
    counter <- counter + 1
    
    print(paste("I just got better:", counter, sep = " "))
    
  }
  
}
```

```{r}
# Some cleaning
rm(new_data, random_order, result_int, variable_cat, variable_num)
```

## Best matching

### Prepare the best matching table
```{r}
best_match <- start_data[permutation$orders, ]
```

Just some table factorization
```{r}
best_match$new_group <- as.factor(new_groups)

best_match$random_group <- as.factor(sample(best_match$new_group, dim(start_data), replace = FALSE))

best_match$Pool <- as.factor(best_match$Pool)

best_match$Grup <- as.factor(best_match$Grup)
```

### Inspecting the table

#### Numeric variables

#### Age
```{r}
ggplot(best_match, aes(x = new_group, y = Day, fill = new_group)) +
  geom_boxplot(alpha = 1) + 
  scale_fill_manual(values = my_colors) + 
  geom_hline(yintercept = mean(best_match$Day), linetype = "dashed") +
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Permutation result")

ggplot(best_match, aes(x = Pool, y = Day, fill = Pool)) +
  geom_boxplot(alpha = .8) + 
  scale_fill_manual(values = my_colors) + 
  geom_hline(yintercept = mean(best_match$Day), linetype = "dashed") + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Manual result")

ggplot(best_match, aes(x = random_group, y = Day, fill = random_group)) +
  geom_boxplot(alpha = .6) + 
  scale_fill_manual(values = my_colors) + 
  geom_hline(yintercept = mean(best_match$Day), linetype = "dashed") + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Random result")
```

#### Discrete variables

#### Sex
```{r}
ggplot(best_match, aes(x = new_group, fill = Sex)) +
  geom_bar(alpha = 1) + 
  scale_fill_manual(values = my_colors) + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Permutation result")

ggplot(best_match, aes(x = Pool, fill = Sex)) +
  geom_bar(alpha = .8) + 
  scale_fill_manual(values = my_colors) + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Manual result")

ggplot(best_match, aes(x = random_group, fill = Sex)) +
  geom_bar(alpha = .6) + 
  scale_fill_manual(values = my_colors) + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Random result")
```

#### Clinical Group
```{r}
ggplot(best_match, aes(x = new_group, fill = Grup)) +
  geom_bar(alpha = 1) + 
  scale_fill_manual(values = my_colors) + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Permutation result")

ggplot(best_match, aes(x = Pool, fill = Grup)) +
  geom_bar(alpha = .8) + 
  scale_fill_manual(values = my_colors) + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Manual result")

ggplot(best_match, aes(x = random_group, fill = Grup)) +
  geom_bar(alpha = .6) + 
  scale_fill_manual(values = my_colors) + 
  theme_bw() + theme(aspect.ratio = 1/2) + ggtitle("Random result")
```

# Session Info
```{r}
info <- sessionInfo()

info
```

# Save Env
```{r}
# save.image(file = paste("./", Sys.Date(), "_matching_clinicaldata.RData", sep = ""))
```
