---
title: "randomization and plate design script (bulk) RNAseq"
author: "Lorenzo Bonaguro & Rainer Knoll"
date: "05 3 2024"
output: html_document
---

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 0. Introduction

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------


This script can be used to randomize a data table for a selection of metadata information. This is especially useful for large bulk RNAseq experiments and reduces batch effects. In addition to the randomization, samples are also ordered onto a 96-well plate. 

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 1. packages and functions

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

## functions

functions for this script are stored in the *function_randomization_plate_design.R* script. 

```{r}
source(file = "./source/functions_randomization_plate_design.R")
```

## packages

```{r}
library(ggplot2)
library(readr)
library(readxl)
library(openxlsx)
```


# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 2. data preparation

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

Load your dataframe with the information for randomization into R as *input_data*. Here, we load an example dataset, which contains the information about:

- donor
- center
- sex
- age

The variables center, sex and age should be randomized to avoid batch effects. 

```{r}
input_data <- read.csv("./source/example_data.csv")
input_data
```

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 3. randomization

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

## performing the randomization

The randomization is performed with the *randomize_data* function. Please carefully select your parameters for randomization. 

- input: your input data
- n_samples_per_batch: number of samples in each batch. E.g. 24 samples per RNA extraction batch is common. 
- n_batches: amount of extraction rounds (if empty: default is the rounded result from total n samples divided by n_samples_per_batch)
- randomize_numeric: numeric metadata to randomize, must be a column in input.
- randomize_categorical: categorical metadata to randomize, must be a column in input.
- reproducible: if TRUE, enables reproducibility using seed.
- n_perm: numeric input, if not specified automated best selection.

```{r}
random_data <- randomize_data(input = input_data,
                               n_samples_per_batch = 24,
                               n_batches = 8, # nrow(input_data)/n_samples_per_batch, if NULL calculates automatically
                               randomize_numeric = c("age"),
                               randomize_categorical = c("center", "sex"),
                               reproducible = T, # if FALSE, the function will perform differently every time (number of perumations required)
                               n_perm = NULL) # NULL = automatic, else use numeric value for number of permutations, e.g. 1e+06
```

the final output of the function will contain the data in randomized order with annotation of the batches with n_samples_per_batch as previously selected.

```{r}
random_data
```

## visualizing results

TO visualize the randomized results, some plots are useful. We look at the distribution of the previously selected variables in the randomized data.

If you are unhappy, you can try to increase the number of permutations. 

### numeric variables

```{r}
plot_numeric(random_data = random_data, randomize_numeric = "age")
```

### categorical variables

```{r}
plot_categorical(random_data = random_data, randomize_categorical = "sex")
plot_categorical(random_data = random_data, randomize_categorical = "center")
```

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# 4. plate design (optional)

# -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

We nicely randomized our dataset, which can be sufficient for some purposes. In the case of bulk RNAseq, it would be nice to have 96-well plate layout, as the samples will be processed in that manner. This helps the handling person in the lab and gives a nice overview.

Please choose which labels should be put on the 96-well plate distribution. E.g. randomized order and donor ID are important. After randomization, a LAB ID might be generated which should be added to the dataframe and is quite useful in the 96-well plate overview. 

Here, a random LAB ID is generated (usually they have 5 digits).

```{r}
# add LAB ID to dataset (multiple options, e.g. read new dataframe, match, merge, ...)
random_data$LAB_ID <- seq(30000, 30000-1+nrow(random_data), 1)
```

An excel sheet in 96-well plate format will be created using the *create_plate_design* function. Please state which variables should be listed on the plate, here we use the randomized_order, donor and PRECISE_ID, which will look e.g. like this:
1 donor_38 30000. Please also state the path/name of your xlsx file. 

The resulting xlsx file has the plate in the respective sheets and is colored by 24-batches (column-wise).

```{r}
create_plate_design(random_data = random_data, 
                    variables_to_include = c("randomized_order", "donor", "LAB_ID"), 
                    path = "plate_design.xlsx")
```






